---
title: "Accounting for migration in spatiotemporal models of county-level health"
date: ""
author: "Hannah Olson-Williams, Amy Cochran"
format: docx  
editor: visual
bibliography: references.bib
---

<!--# perhaps should add: -->

<!--# some explanation of spatial error versus spatial lag -->

<!--# packages used to assess each aim -->

<!--#  need sensitivity analysis with natality before adding urb; so basically 1) add mig; 2) add natality; 3) decide about natality; 3) add urb  -->

### [Approach]{.underline}

We test the following hypotheses:

1.  County-to-county migration flows improve the explainability of spatiotemporal models of county-level health outcomes.

2.  County-to-county migration flows improve the explainability of spatiotemporal models of county-level health outcomes, even after accounting for rurality.

3.  Taking into account unmeasured factors in county-to-county migration flows improves our ability to explain county-level health outcomes as well as the differential role that migration plays in urban versus rural counties.

Considering the exploratory nature of these hypotheses, we first establish a baseline model which is similar to autoregressive county-level mortality models used in the past [@yang2012]. This baseline model includes yearly mortality rates as a function of time (treated categorically) and premature mortality rates lagged by one year. Random intercepts for each destination county are included to account for variation between counties, following best practies for ecological studies[@greenland1989]. We tested splines with two to five degrees of freedom to account for potential nonlinearity in mortality rates. The baseline model (without splines) is shown below:

$$  y_{it} = \beta_0 + \beta_{1}t_{13i} + \beta_{2}t_{14i} + \beta_{3}t_{15i}  + \beta_{4}t_{16i} + \beta_{5}t_{17i} + \beta_{6}t_{18i} + \beta_{7}t_{19i} + \beta_{8} y_{t-1,i} + \upsilon_{i}t + \mu_{i} + \epsilon_{it}    $$

where\
$y_{it}$ represents the predicted age-adjusted mortality rate of each destination county $i$ at year $t$.\
$t_{yi}$ are binary factor variables for each year of available data.\
$t$ is time, in years, treated continuously.\
$\beta_j$ are unknown regression coefficients.\
$y_{t-1,i}$ is the lagged age-adjusted mortality rate for each destination county $i$ at year $t-1$.\
$\mu_{i}$ is a random intercept for each destination county $i$.

$\epsilon_{it}$ represents an error term for each destination county $i$ at year $t$.

\

We began with our baseline model and iteratively added splines for the prior year's average county-level premature mortality rate and the migration term to capture potential nonlinear trends. To determine the best fitting model, we analyzed several models using the Bayesian Information Criterion (BIC), which is commonly used to identify models with the greatest explanatory power \[\@kuha2004\]. We observed that keeping the degrees of freedom equal for the splines of both the migration term and the premature mortality rate resulted in lower BIC scores, indicating better model fit. All models accounted for both spatial and random error and were created using the `splm` and `splines` R packages \[\@splines\]\[\@millo2023\].

To determine whether county-to-county migration flows improve the explainability of spatiotemporal models of county-level health outcomes (Hypothesis 1), we compared the baseline model to models that account for migration. To quantify migration, we developed a novel "migration term" which accounts for the mortality rates of individuals who have moved to a county of interest during a given period. This term is essentially a weighted average of the mortality rates of all the origin counties. We calculated the following for each change in year and for each destination county $i$:

$$  mig_{it} = \frac{ \sum_{j\ne i} out_{jit} y_{j,t-1} + y_{i,t-1} (pop_{i, t-1} - \sum_{j\ne i} out_{ijt})}{ \sum_{j\ne i} out_{jit} + (pop_{i,t-1} - \sum_{j\ne i} out_{ijt})} $$

where

$y_{it}$ was defined above\
$out_{ijt}$ represents the number of migrants from a unique origin county $j$ who migrated to a destination county $i$ between year $t-1$ and year $t$.\
$pop_{i, t-1}$ is the population of an unique destination county $i$ at initial year $t-1$\

The table below presents descriptive statistics for $y_{it}$​ (premature age-adjusted mortality rate) and migration rates, categorized by year and county type. Note that migration rates for the year 2011 are not included. This is because the migration-flow data starts from 2011 to 2012, and these values are shown in the 2012 section of the table, corresponding with the 2012 premature age-adjusted mortality rates.

```{r table1 as dreamed up by chatgpt }

# probably need to find a way to add units to this somehow 

load("data_processed/migterm_imp.Rdata")

migterm_imp$migrat = as.numeric(migterm_imp$totin_d)/as.numeric(migterm_imp$pop_d0)*1000

migterm_imp = migterm_imp %>% 
   mutate(
    rural = factor(rural, levels = c(0, 1), labels = c("Urban", "Rural"))
  )

# Define labels for the table
table1::label(migterm_imp$destid) = "County"
table1::label(migterm_imp$year) = "Year"
table1::label(migterm_imp$migrat) = "Migration rate"
table1::label(migterm_imp$rate_d1) = "Premature age-adjusted mortality rate"
table1::label(migterm_imp$rural) = "Urban/Rural"

# Define units for the columns
table1::units(migterm_imp$migrat) <- "per 1,000"
table1::units(migterm_imp$rate_d1) <- "per 100,000"

# Filter data for rural
rural_data <- migterm_imp %>% filter(rural == "Rural")

# Filter data for urban
urban_data <- migterm_imp %>% filter(rural == "Urban")

# Generate the table for rural with title
table_rural <- table1::table1(~ migrat + rate_d1 | year, 
                              data = rural_data, 
                              overall = "Total", 
                              title = "Table 1: Rural Counties",
                              caption = "Migration Rate and Mortality Rate stratified by Year for Rural Counties")

# Generate the table for urban with title
table_urban <- table1::table1(~ migrat + rate_d1 | year, 
                              data = urban_data, 
                              overall = "Total", 
                              title = "Table 2: Urban Counties",
                              caption = "Migration Rate and Mortality Rate stratified by Year for Urban Counties")

# Display the tables
table_rural
table_urban

```

Our weighted-average approach implicitly accounts for between-county selection for migration but does not account for within-county selection of migrants. In other words, our best guess for the mortality rate of migrants is exactly the mortality rate of their original county; meanwhile, some counties are more likely to be destinations for migration than others.

To determine whether the role that county-to-county migration flows play in county-level health outcomes differs significantly between rural and urban counties (Hypothesis 2), we added a factor term for urbanicity including interaction terms between urbanicity and $y_{i,t-1}$ and urbanicity and $mig_{i,t}$ to the models accounting for spatial error used to test Hypothesis 1. We again compared these models using BIC while iteratively adding splines with equal degrees of freedom for migration and lagged premature mortality.

To maximize the number of counties per category, we chose to use only two urbanization categorizations: urban and rural, where urban includes large central metro, medium metro, and small metro counties, and rural includes micropolitan and noncore counties. We assessed \`r length(unique(migterm_imp$destid[migterm_imp$rural == 1\])) \` unique rural counties and r length(unique(migterm_imp$destid[migterm_imp$rural == 0\])) unique urban counties. We utilized the same steps outlined above to determine which model had the most explanatory power and lowest BIC score.

Finally, to determine whether taking into account unmeasured factors in county-to-county migration flows improves our ability to explain county-level health outcomes (Hypothesis 3), we added parameters $k_{ij}$ and $l_i$ to the premature mortality rate of each origin $j$ and each destination $i$ within our equation for the migration term. We call this new simulated migration term $smig_{it}$. The equation for $smig_{it}$ is shown below:

$$  smig_{it}(k_{ij}, l_i) = \frac{ \sum_{j\ne i} out_{jit} ( y_{j,t-1} + k_{ij}) + (y_{i,t-1} + l_i) (pop_{i, t-1} - \sum_{j\ne i} out_{ijt})}{ \sum_{j\ne i} out_{jit} + (pop_{i,t-1} - \sum_{j\ne i} out_{ijt})} $$

We use $smig_{it}$ to simulate within-county selection for migration: when the parameter $k_{ij}$ is subtracted (ie decreased mortality rate), we simulate migration by individuals who are healthier than the average of their origin $j$. When the parameter $k_{ij}$ is added (ie increased mortality rate), we simulate migration by individuals who are unhealthier than the average of their origin $j$. Likewise, when the parameter $l_i$ is subtracted (ie decreased mortality rate), we simulate migration by individuals who are healthier than the average of their destination $i$. When the parameter $l_i$ is added (i.e., increased mortality rate), we simulate migration by individuals who are unhealthier than the average of their origin $i$. We specify that $k_{ij}$ depends on both origin $j$ and destination $i$ since health related selection for migration is connected to both the origins and the destinations of counties; meanwhile $l_i$ is dependent only upon the health of destination $i$.

Since we are most interested in how the values of $k_{ij}$ and $l_i$ are related to each other, we chose to simplify $smig_{it}$ such that $l_i$ is held constant and only $k_{ij}$ fluctuates. Therefore, we assign the parameter $d_{ij}$ to be the difference between $k_{ij}$ and $l_i$, thus creating a new equation for $k_{ij}$ in terms of $l_i$ and $d_{ij}$ such that $k_{ij} = l_i+ d_{ij}$. The value of $d_{i,j}$ can be thought of as a modeling term to explain health-related selection bias. <!--# amy's phrasing: "refer to the dij as a modeling term that models health related selection bias" -->Then our equation for $smig_{it}(k_{ij})$ becomes:

$$  smig_{it}(l_i+d_{ij}, l_i) = \frac{ \sum_{j\ne i} out_{jit} ( y_{j,t-1} + l_i + d_{ij}) + (y_{i,t-1} +l_i)( pop_{i, t-1} - \sum_{j\ne i} out_{ijt})}{ \sum_{j\ne i} out_{jit} + (pop_{i,t-1} - \sum_{j\ne i} out_{ijt})} $$

which can be simplified into the following:

$$  smig_{it}(d_{ij}) = \frac{ \sum_{j\ne i} out_{jit} ( y_{j,t-1} + d_{ij}) + y_{i,t-1} (pop_{i, t-1} - \sum_{j\ne i} out_{ijt})}{ \sum_{j\ne i} out_{jit} + (pop_{i,t-1} - \sum_{j\ne i} out_{ijt})} +l_i$$

Therefore, using the equation above, we simulate health-related selection occurring as a result of the differences in health between the origin and the destination counties.

To determine which values of $d_{ij}$ best explain trends in county-level premature mortality rates, we replace $mig_{it}$ with $smig_{it}(d_{ij})$ in the model with the most explanatory power as determined by our previous hypotheses.

We then test potential values of $d_{ij}$ to determine which value produces the lowest BIC score. We incremented by 50 from -200 to 200 such that migrants had an estimated premature mortality rate that was between 200 premature deaths per 100,000 population higher or 200 premature deaths per 100,000 population higher lower than their origin county.

To understand this phenomenon further, we conducted a grid search to test potential values of $d_{ij}$ when accounting for In a secondary simulation, we varied \$d\_{ij}\$ for each possible combination of origin and destination type. In other words, \$d\_{ij}\$ was different for each migration scenario: rural to rural, urban to rural, rural to urban, and urban to urban. urbanicity, where these values depended on whether origins and destinations were rural or urban. We started with a large increment of 50, ranging from -200 to 200 premature deaths per 100,000 higher than the destination county. To obtain more precise results, we then narrowed our increment to 10 and conducted the same analyses within a smaller range of -40 to 40 premature deaths per 100,000 higher than the destination county.

<!--# present all results for transparency -->

<!--# add colors to plots for other varied -->

<!--# appendix: add plots for every possibilities of dij -->

```{r plot for ksim with urb/rural differences ; echo = FALSE}
load("data_processed/bickdf.Rdata")



bicmin = bickdf %>% filter(kuu == 0 & krr == 0)
#lattice::wireframe(bicspat ~ as.numeric(kru) * as.numeric(kur) , data = bicmin)
ggplot(aes(x = as.numeric(kur), y = as.numeric(bicspat)), data = bicmin) + 
  geom_point() + 
  xlab("Difference in average premature mortality rate\n between urban origin counties and rural destination counties") +
  ylab("BIC") + 
  theme_bw() +
  labs(subtitle = expression(paste("Relationship between BIC and ", d[ij], "\n (assuming that ", d[rr], " = 0, and ", d[uu], " = 0)")),
       title = "Urban-Rural Migration")


ggplot(aes(x = as.numeric(kru), y = as.numeric(bicspat)), data = bicmin) + 
  geom_point() + 
  xlab("Difference in average premature mortality rate\n between rural origin counties and urban destination counties") +
  ylab("BIC") + 
  theme_bw() +
  labs(subtitle = expression(paste("Relationship between BIC and ", d[ij], "\n (assuming that ", d[rr], " = 0, and ", d[uu], " = 0)")),
       title = "Rural-Urban Migration")

```

### [**Results and Conclusions**]{.underline}

To justify further investigation, we first determined that our migration term added explainability (as measured by lower BIC scores) to our most basic autoregressive model accounting for spatial error. The best fitting model in terms of BIC score included splines with four degrees of freedom for both the autoregressive term and our weighted average migration term. This indicates that models incorporating migration had more explanatory power than models that included only the prior year's average county-level mortality rate.

Next, we sought to determine whether our migration term would improve the explainability of our spatiotemporal model, even after accounting for urbanicity. We added interaction terms between urbanicity and lagged premature mortality and urbanicity and migration. After adjusting for urbanicity, we found that the models with the lowest BIC scores included splines with four degrees of freedom each for the migration term and lagged age-adjusted mortality. This further confirmed that the migration term added explainability to our models, even when accounting for urbanicity.

The plot below demonstrates lower BIC scores for models that include the migration term along with an interaction between $y_{it}$ and urbanicity. The difference in BIC scores between models that include the migration term and those that do not is more pronounced when an interaction term for urbanicity is added. This suggests that the migration term significantly enhances model explainability when accounting for urbanicity, indicating that the relationship between average county-level health and migration may differ substantially across urbanicity levels.

```{r spatial models only, echo = FALSE, warning = FALSE, message = FALSE}

spatonly = read.delim("chtc_outputs/modselection_nat.txt", sep = ",")  
ref = spatonly$BIC[spatonly$Model == "fnom"] 

spatonly$refbic = round(spatonly$BIC - ref, 2)   



tf <- spatonly %>%
  mutate(
    `Migration Included` = !grepl("_nomig|nom", Model),
    `Degrees of Freedom` = str_extract_all(Model, "\\d+"),
    `BIC score relative to reference` = refbic,
    `Urbanicity interaction` = grepl("r", Model)
  )
# Convert the list of numbers to a single string or set to "1" if no number is found
tf$`Degrees of Freedom` <- sapply(tf$`Degrees of Freedom`, function(x) {
  if (length(x) == 0) {
    return("1")
  } else {
    return(paste(x, collapse = " "))
  }
})

tf = tf %>% select(`Migration Included`, `Degrees of Freedom`, `BIC score relative to reference`, `Urbanicity interaction`)


tf$`BIC score relative to reference`[tf$`BIC score relative to reference` == 0] = "REF"


tf = tf %>% arrange(`Degrees of Freedom`, `Migration Included`, `Urbanicity interaction`)

ggplot(data = tf, aes(x = `Degrees of Freedom`, 
                      y = as.numeric(`BIC score relative to reference`), 
                      color = `Migration Included`,
                      size = `Urbanicity interaction`,
                      label = ifelse(`BIC score relative to reference` == "REF", "REF", ""))) +    
    geom_point(alpha = 0.6) +  # Include alpha within geom_point
    geom_text(color = "black", size = 5, fontface = "bold", 
              nudge_y = 0.05, check_overlap = TRUE) +  # Add text labels with adjustments
    theme_bw() +
    labs(x = "DF", y = "BIC relative to reference",
         color = "", size = ""
    ) +
    scale_color_manual(
        values = c("TRUE" = "darkblue", "FALSE" = "darkred"),
        labels = c("TRUE" = "Including migration term", "FALSE" = "Excluding migration term"),
        guide = guide_legend(title = NULL)
    ) +
    scale_size_manual(
        values = c("TRUE" = 5, "FALSE" = 2),
        labels = c("TRUE" = "Including urbanicity interaction", "FALSE" = "Excluding urbanicity interaction"),
        guide = guide_legend(title = NULL)
    ) +
    guides(
        color = guide_legend(order = 1), 
        shape = guide_legend(order = 2), 
        size = guide_legend(order = 2),
        alpha = "none"
    ) +
    coord_cartesian(ylim = c(min(as.numeric(tf$`BIC score relative to reference`), na.rm = TRUE) * 1.1, 0)) 

```

Finally, we simulated health-related selection for migration using $d_{ij}$, a modeling term used to measure the difference between the average health of an origin county and the average health of a destination county. When our simulation did not account for the urbanicity of destination and origin counties, the model with the most explanatory power had $d_{ij} = 0$ . As the magnitude of the difference between the health of origin and destination increased, our models had less explanatory power and higher BIC scores. This is shown in the plot below.

```{r ksim without urb, just k, echo = FALSE}
load("data_processed/bick_nourb_tot.Rdata")

ggplot(data = bick_nourb_tot, aes(x = as.numeric(newk), y = as.numeric(bicspat))) +
  geom_point() +
  theme_bw() + 
  labs(x = expression(d[ij]), 
       y = "BIC score")
```

In a secondary simulation, we varied \$d\_{ij}\$ for each possible combination of origin and destination type. In other words, \$d\_{ij}\$ was different for each migration scenario: rural to rural, urban to rural, rural to urban, and urban to urban. We found no evidence of health-related selection for migration since the model with the lowest BIC score included $d_{ij} =0$ for every origin-destination pair.

Even when moving from counties that differ in urbanicity, health does not appear to be a factor in deciding who migrates and who does not. This is somewhat counter to what the literature says about health and migration. Typically migrants are healthier on average than the general population. <!--# cite --> But it is important to consider relative to whom. In our models, we have found that migrants with rural destinations are no more or less healthy than others within their urban origin counties - this could indicate that older (less healthy) individuals tend to move to rural places from urban places. Meanwhile, we found that migrants with urban destinations tend to be slightly less healthy than others within their rural origin counties - this could indicate that people migrate to cities seeking resources which could be tied to health CITE. Interestingly, when we isolate the effects of rural-urban migration and urban-rural migration on BIC score separately, we see that trends in BIC score are primarily driven by rural-urban migration; there does not appear to be a strong relationship between BIC score and the difference in premature mortality rate of urban origin and rural destination counties, but there does appear to be a strong relationship between BIC score and the difference in premature mortality rate of rural origin and urban destination counties.

```{r lotsa plots to see if there are any trends }
load("data_processed/bickdf.Rdata")

bicmin = bickdf %>% filter(kur == 0 & kru == 0)
bicmin$colorvar = ifelse(bicmin$krr >0, 1, 0)

lattice::wireframe(bicspat ~ as.numeric(kuu) * as.numeric(krr) , data = bicmin)


ggplot(aes(x = as.numeric(kuu), y = as.numeric(bicspat), color = as.factor(colorvar)), data = bicmin) +
  geom_point() +
  xlab("Difference in average premature mortality rate\n between urban origin counties and rural destination counties") +
  ylab("BIC") +
  theme_bw() +
  labs(subtitle = expression(paste("Relationship between BIC and ", d[ij], "\n (assuming that ", d[rr], " = 0, and ", d[uu], " = 0)")),
       title = "Urban-Rural Migration") +
  scale_color_manual(
    values = c("0" = "red", "1" = "blue"),
    labels = c("0" = expression(d[rr] <= 0), 
               "1" = expression(d[rr] >0)))



```

```{r plot for ksim_urbrural }
load("data_processed/bickdf.Rdata")

bicmin = bickdf %>% filter(kuu == 0 & krr == 0)

bicmin$colorvar = ifelse(bicmin$kur >0, 1, 0)


lattice::wireframe(bicspat ~ as.numeric(kru) * as.numeric(kur) , data = bicmin)
ggplot(aes(x = as.numeric(kur), y = as.numeric(bicspat)), data = bicmin) +
  geom_point() +
  xlab("Difference in average premature mortality rate\n between urban origin counties and rural destination counties") +
  ylab("BIC") +
  theme_bw() +
  labs(subtitle = expression(paste("Relationship between BIC and ", d[ij], "\n (assuming that ", d[rr], " = 0, and ", d[uu], " = 0)")),
       title = "Urban-Rural Migration")



ggplot(aes(x = as.numeric(kru), y = as.numeric(bicspat), color = as.factor(colorvar)), data = bicmin) + 
  geom_point() + 
  xlab("Difference in average premature mortality rate\n between rural origin counties and urban destination counties") +
  ylab("BIC") + 
  theme_bw() +
  labs(subtitle = expression(paste("Relationship between BIC and ", d[ij], "\n (assuming that ", d[rr], " = 0, and ", d[uu], " = 0)")),
       title = "Rural-Urban Migration",
       color = "Modeling term for selection from urban to rural") +
  scale_color_manual(
    values = c("0" = "red", "1" = "blue"),
    labels = c("0" = expression(d[ur] <= 0), 
               "1" = expression(d[ur] >0)))

```

```{r}
load("data_processed/bick_restrict.Rdata")

bicmin = bick_restrict %>% filter(kuu == 0 & krr == 0)
bicmin$colorvar = ifelse(bicmin$kur >0, 1, 0)

lattice::wireframe(bicspat ~ as.numeric(kur) * as.numeric(kru) , data = bicmin)


ggplot(aes(x = as.numeric(kru), y = as.numeric(bicspat), color = as.factor(colorvar)), data = bicmin) +
  geom_point() +
  xlab("Difference in average premature mortality rate\n between urban origin counties and rural destination counties") +
  ylab("BIC") +
  theme_bw() +
  labs(subtitle = expression(paste("Relationship between BIC and ", d[ij], "\n (assuming that ", d[rr], " = 0, and ", d[uu], " = 0)")),
       title = "Urban-Rural Migration") +
  scale_color_manual(
    values = c("0" = "red", "1" = "blue"),
    labels = c("0" = expression(d[rr] <= 0), 
               "1" = expression(d[rr] >0)))
```

### Discussion and Limitations {.underline}

We consider several pitfalls and alternatives. First, because our migration term is essentially a weighted average of the mortality rates of origin counties, the migration term and lagged mortality rates are highly correlated, resulting in multicollinearity. As a result, we cannot easily interpret the effect of the migration term itself, other than knowing that its inclusion results in a model with more explanatory power over all. Application of results must consider this shortcoming.

Second, we selected our outcome of interest, county-level premature age-adjusted mortality, because it is comparable across years and locations and because it is consistently and robustly defined. However, in 2019, the median age of all movers across the US was under 30 years old [@s0701:g] - therefore, even when accounting for premature mortality (deaths occurring in individuals under age 75), the population that is migrating is not necessarily the same population that is dying. Subsequently, trends in county-level premature mortality might be better explained by trends in county-level migration twenty to fifty years earlier. That said, county-level measures of premature mortality are commonly used as a proxy for county overall health. For instance, a 1983 report by the WHO recommending measures in mortality and morbidity to understand the relationship between health and migration [@gushulak2006], and measures of length of life are allocated 50% weight in the County Health Rankings Model of place-based health outcomes [@explore]. Ideally we would use a more precise age-group-specific measure of mortality paired with an age-group-specific measure of migration; however, this would require a migration data source other than the IRS migration flow data which we use here and therefore could not include analysis of the entire United States. Additionally, ideally, we would confirm our findings by assessing the health of migrants themselves. This task would require individual longitudinal data. This is not currently within the scope of this project.

One final limitation of IRS data (which we are attempting to leverage as a strength in our analyses) is that households that do not file tax returns cannot be included in migration estimates. Therefore, many university students, low-income households, and workers who receive informal wages are systematically missing from IRS migration estimates[@dewaard2022]. These individuals may be more likely to be included in ACS and Census estimates [@bureau] though no population estimate is perfect. Because IRS county-to-county migration data includes only individuals who file taxes with the US government both before and after they migrate, we assume that individuals represented in the IRS migration flows data are less likely to be pushed to migrate by potentially health-related factors such as violence, famine, and corruption than individuals who migrate but are excluded from the IRS data. Therefore, any differences in changes in health between counties may be the result of self-selection by IRS-represented migrants or health-related selection by nonIRS-represented migrants. Despite these limitations, IRS migration data has been used many times in the past to complete complex and accurate analyses of US migration patterns. For instance, IRS migration flow data has been used to estimate the effects of sea-level rise on geographic distribution of the US population[@hauer2017], measure recovery after Hurricanes Katrina and Rita[@curtis2015], and assess the economic impacts of migration resulting from environmental hazards [@shumway2014].

<!--# paragraph in discussion section about policy lens; NOTE: all data is pre covid -->

Additionally, since we rely on IRS migration flow data from 2011 through 2019, all data are pre-COVID. The COVID pandemic caused XXX migration during the time period and may have also influenced some individuals to migrate from rural to urban and from urban to rural locations more than others. CITE. <!--# policy lens -->

There are many factors that contribute to the health of a place. In our analyses, we have chosen to emphasize the potential impact of migration or movement between places on place-based health. We accounted for intercounty selection for migration (ie some counties are more likely to experience migration than others) using our weighted average migration term and intracounty selection for migration (ie some individuals are more likely to migrate than others) using our $k_i$ and $k_j$ simulation. Future research should investigate a potential relationship between $k_i$ and $k_j$ - perhaps some destinations are more likely to attract migrants of relative health than others.

We cannot fully understand county-level health without first understanding county interconnectedness and how it drives place-based health disparities. Recognizing that the United States is heterogeneous and that the mechanisms by which migration may impact health may also be heterogeneous, we attempt to quantify rural-urban differences in the relationship between migration and place-based health. The long-term goal of this work is to contribute to understanding the mechanisms by which mobility may be related to place-based health disparities so that local decision-makers may account for patterns in mobility when creating policy towards improved health for all people in all places. Understanding rural-urban mobility patterns is a necessary first step towards understanding rural-urban health disparities.
